Translation of B Implementations to the LLVM: Specification

### 2014-05-13 DOCUMENTATION/DerivedTypes.tex
Adding declaration for Structures and Arrays

\begin{pascalcode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance -1;
\end{pascalcode}

@@

\begin{align*}
\begin{split}
  \trad{n}{\B{RecField}} \isdef \\
   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

@@

%TODO Attributions and initializations

@@

The arrays in B0 are represented by functions where an index points to an 
element. By default, the element of index not initialized has a zero value.

This section shows the relation between arrays B0 and LLVM. The sequence of
elements are defining an array, getting an element and setting an
element. The white code block and gray code block are respectively B code and LLVM code.

\subsubsection{Declaring an array}

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1,...,in) are defined according 
to the formula:\\
 $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - e.in\_start+1)$.

The following code declares a multidimensional global array called ``Array2'' 
with 16 (2*2*4) integer numbers with 32 bits.

\begin{llvmcode}
@array2 = internal global [2 x [2 x [4 x i32]]]
\end{llvmcode}

@@

\begin{pascalcode}
array2(0,0,0) := 5;
// A sequence of attributions ...
array2(1,1,3) := 5;
\end{pascalcode}

@@

\begin{pascalcode}
 array2 := ( (0..1) * (0..1) * (0..3) * {1});
 array2 := { 0|->0|->0|->2 , 0|->0|->1|->(a3d(0,0,0)+1), ...  };
\end{pascalcode}

@@

\begin{pascalcode} 
VAR xx, yy, zz IN
	xx=0;
	yy=1;
	zz=3;
	array2(x,y,z):=7
END
\end{pascalcode}

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 0, i32* %xx
store i32 1, i32* %yy
store i32 3, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

@@

\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indexes that 
calculates the address of element. This rule also supports to translate 
expressions of arrays.

@@

\begin{align*}
&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\ \\
&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \PH{\ell} = \lbl \IN \\
&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
&\quad \ELSE \LET  p' \sep v \sep v' \sep t'  = \trad{t,v,itype,istart}{Cast} \IN \\
&\quad \llvm{",\PH{t} \PH{\ell} "} \sep  \llvm{"\PH{p} \PH{p'}  \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
\end{align*}

@@

\paragraph{Restriction on array:} The translation of direct attribution
 between arrays is not yet supported, for example:

\begin{pascalcode}
 array1 := array2;
\end{pascalcode}

%TODO Describe here the restrictions of declarations arrays
\paragraph{Relevant restrictions of work:}
\begin{itemize}
\item  B0 definition - An array must be initialized declaring all values 
to avoid no determinism.
\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:

>>>>

Initializing a record with a sequence of expressions:
\[
  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
\]

Setting a record field to a variable:
\[
  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
\]

A B code example:

\begin{pascalcode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance-1;
\end{pascalcode}

@@

\begin{align*}
\begin{split}
  \trad{n}{\B{RecField}} \isdef \\
   &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
   &code \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
   &\IF \B{n.base.type} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
   &\quad \THEN  code \sep  \ell  \sep \SBType(\B{n})\\
   &\IF \B{n.base.type} \mbox{ is \B{RecField} }  \THEN \\
   &\LET \ell2 = \lbl  \AND  t = n.type \AND v = \B{tail}(n.element) \AND \\
   &\quad n2.base  \BE ?? \AND n2.element = \B{head}(n.element) \AND a, b, c = \trad{n2}{RecField}
\IN\\
   &\quad  code \conc  \llvm{"\PH{\ell2} = getelementptr \PH{t}* \PH{\ell}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad \sep  \ell2  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

@@

%[[ Write a text here to define the rule for it]]

There are three attributions cases related to array.
\begin{description}

\item[Case 1.1] - Initializing an unique value to an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2 \times 0..3  }_{seq\ Exp} \times \underbrace{\{5\}}_{Exp} }_{Product}\}}_{Array}
\]
\item[Case 1.2] - Initializing different values to an array\footnote{B0 definition requires the declaration of all values.}:
\[
\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{5}_{Exp} , ... ,
(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{5}_{Exp} }_{seq\ Map}\}}_{Array}
\]

Theses cases use a declared constant that is attributed to the variable:
\begin{llvmcode}
@array = global [3 x [3 x [4 x i32]]] 
[[3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
 	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
\end{llvmcode}

The rule for declare array constants:

\begin{align*}
\begin{split}
  \trad{array}{\B{Type}} \isdef \\
   & \LET \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}\\
   & \quad \ELSIF array \mbox{ is \B{Map} } \THEN\\
   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
   & \quad \IN \llvm{" global \PH{dec}"}\\
\end{split}
\end{align*}
\VGM{Alguma sugestão com relação as funções:head, tail.}

\begin{align*}
\begin{split}
  \trad{type, seqIntl, value}{\B{Product}} \isdef \\
   &\LET \B{h = head(seqIntl)} \AND\\
   & \B{times = h.end - h.start + 1} \AND\\
   & values = \trad{times,type,value}{\B{TV}} \IN\\
   & \IF \B{seqIntl.length} == 1 \THEN  \\
   & \llvm{"[\PH{times} x \PH{type}] [\PH{values}]"}  \\
   & \ELSE \LET \\
   &  tvalues = \trad{type,tail(seqIntl),values}{\B{Product}} \\
   & \IN \\
   & \llvm{"[\PH{times} x \PH{type}]  \PH{tvalues}"}  \\
\end{split}
\end{align*}

\begin{align*}
\begin{split}
  \trad{times, type, value}{\B{TV}} \isdef \\
   &\LET res1 = \llvm{" \PH{type} \PH{value} "} \IN\\
   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
   & \ELSE \LET   res2 =   \trad{times-1,type,value}{\B{TV}}  \IN  \\
   & \llvm{"\PH{res1} , \PH{res2}"} \\
\end{split}
\end{align*}

The ``head'' function returns the first element from sequence.
The ``tail'' function returns the given sequence without its first element.

\begin{comment}
\$VarName = global [size X type] X [(type value,)^+]

|array|Type = LET
				  IF array is Product THEN dec = |array.arg1, array.arg2 |CProduct  ( Quebrando o padrão?)
				  ELSEIF array is seq Map THEN dec = |array.Map |Map
				IN	  
					" global "^ dec
				END
|sIntl, value |CProduct = 
    	IF sIntl is  empty THEN  ""
	ELSE THEN
		LET sizeType = |sIntl, "i32"|PrintSizeType &
		 tailSizeType=|sIntl.tail,"i32"|PrintSizeType & 
		 values= |tailSizeType,values|CProduct IN
		"[sizeType]  [ tailSizeType  values ]"		
		END
	END

|sIntl,t| PrintSizeType =
    	IF sIntl.size == 1  THEN  "t"
	ELSE THEN
		LET size = sIntl.head.size()  &
			type = |sIntl.tail,t| PrintSizeType    &
		"[size X type ]"		
		END
	END

|array|Type = IF t is Map THEN ...
\end{comment}

\item[Case 2.1] - Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]
\item[Case 2.2] - Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem}
\]
\item[Case 3] - Attribution between arrays\footnote{B0 definition requires the same type.}:
\[
\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
\]
\end{description}

@@

The arrays in B0 are represented by functions where an index points to an 
element. By default, the element of index not initialized has a zero value.
The different ways to evaluate an array are described as follow:

@@

\begin{pascalcode}
array2(0,0,0) := 5;
array2(0,0,1) := 5;
// ... - A sequence of attributions
array2(1,1,3) := 5;
\end{pascalcode}

@@

\begin{pascalcode}
 array2 := ( (0..1) * (0..1) * (0..3) * {1});
 array2 := { 0|->0|->0|->2 , 0|->0|->1|->(array2(0,0,0)+1), ...  };
\end{pascalcode}

@@

\begin{pascalcode} 
VAR xx, yy, zz IN
	xx=1;
	yy=1;
	zz=3;
	array2(x,y,z):=7
END
\end{pascalcode}

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 0, 1, 3 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 1, i32* %xx
store i32 1, i32* %yy
store i32 3, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

@@

\paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.

\paragraph{seq Map} declares every address and values from array, it is represented by \textbf{case 1.2}.

\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indexes that 
calculates the address of element. This rule also supports to translate 
expressions of arrays. This rule is used to support array attributions 
from \textbf{cases} \textbf{2.1} and \textbf{2.2}.

@@

\begin{align*}
&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \PH{\ell} = \lbl \IN \\
&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
&\quad \ELSE \LET  p' \sep v \sep v' \sep t'  = \trad{t,v,itype,istart}{Cast} \IN \\
&\quad \llvm{",\PH{t} \PH{\ell} "} \sep  \llvm{"\PH{p} \PH{p'}  \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
\end{align*}

@@

\paragraph{Relevant translation restrictions:}
\begin{itemize}
\item  B0 definition - An array must be initialized declaring all values 
to avoid no determinism.
\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:

### 2014-05-13 DOCUMENTATION/specification.tex
Adding declaration for Structures and Arrays

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage{color}
\usepackage{minted}
\usepackage{array}

@@

\date{March 2014}
\definecolor{bg}{rgb}{0.93,0.93,0.93}
\definecolor{white}{rgb}{1,1,1}
\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=white}

@@

We detail here the constructs that are considered in the compilation to the
LLVM. This description introduces the notations used in the definition of the
translation. The elements of the B language are denoted using a \B{sans serif}
font.

@@

\begin{tabular}[t]{rcl}
	\hline
  	\multicolumn{3}{|c|}{Enumerated set: \B{Enumeration}} \\
        \hline
        \B{elements} & : & seq \B{Enumerated} \\
	\hline
  	\multicolumn{3}{|c|}{Enumerated value: \B{Enumerated}} \\
        \hline
        \B{id} & : & \B{name} \\
        \B{type} & : & \B{Enumeration} \\
	\hline
  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
        \hline
        \B{elements} & : & seq \B{Field} \\
	\hline
  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
        \hline
        \B{id} & : & \B{Name} \\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record: \B{Record}} \\
        \hline
	\B{type} & : & \B{Structure}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
        \hline
	\B{base} & : &  \B{Name} \\
	\B{element} & : & seq \B{Name} \\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The data types \textbf{structures} and \textbf{arrays} require detailed 
representation in the abstract syntax structure from the table \ref{tab:node-attr}.
The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.

@@

\begin{description}

\item[structures] The structure types of B are naturally encoded as structure
  types in LLVM.

Declaring an structure type:
\[
  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
\]
Initializing a record with a sequence of expressions:
\[
  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
\]

Setting a record field to a variable:
\[
  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
\]

\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.
Declaring an array type:
\[
\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
\]

There are three attributions cases related to array.
\subitem\textbf{Case 1.1} - Initializing an unique value to an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2  }_{seq\ Exp} \times \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
\]
\subitem\textbf{Case 1.2} - Initializing different values to an array:
\[
\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(0 \mapsto 0) \mapsto  100,(0 \mapsto 1) \mapsto  101}_{seq\ Map}\}}_{Array}
\]

\subitem\textbf{Case 2.1} - Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]

\subitem\textbf{Case 2.2} - Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem}
\]

\subitem\textbf{Case 3} - Attribution between arrays:
\[
\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
\]
\end{description}

@@

To facilitate the defintiion of rules that apply to similar cases, we introduce
mapping $\opmap$, that associates B operators to corresponding LLVM keywords:
$$\opmap = \{
\begin{array}[t]{l}
  = \mapsto \llvm{"eq"},
  \ne \mapsto \llvm{"ne"},
  < \mapsto \llvm{"slt"},
  \le \mapsto \llvm{"sle"},
  > \mapsto \llvm{"sgt"},
  \ge \mapsto \llvm{"sge"}, \\
  + \mapsto \llvm{"add"} \}.
  \end{array}
$$

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage{color}
\usepackage{verbatim}
\usepackage{minted}
\usepackage{array}

@@

\date{May 2014}
\definecolor{bg}{rgb}{0.93,0.93,0.93}
\definecolor{white}{rgb}{1,1,1}
\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=white}

@@

We detail here the constructs that are considered in the compilation to the
LLVM. This description introduces the notations used in the definition of the
translation. The elements of the B language are denoted using a \B{sans serif}
font, when used the code block so white code block and gray code block means
respectively B code and LLVM code.

@@

\begin{tabular}[t]{rcl}
	\hline
  	\multicolumn{3}{|c|}{Enumerated set: \B{Enumeration}} \\
        \hline
        \B{elements} & : & seq \B{Enumerated} \\
	\hline
  	\multicolumn{3}{|c|}{Enumerated value: \B{Enumerated}} \\
        \hline
        \B{id} & : & \B{name} \\
        \B{type} & : & \B{Enumeration} \\
	\hline
  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
        \hline
        \B{elements} & : & seq \B{Field} \\
	\hline
  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
        \hline
        \B{id} & : & \B{Name} \\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record: \B{Record}} \\
        \hline
	\B{type} & : & \B{Structure}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
        \hline
	\B{base} & : &  \B{Name} \\
	\B{element} & : & seq \B{Name} \\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Interval}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\multicolumn{3}{c}{ \B{start}, \B{end} : \B{Exp} } \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The data types \textbf{structures} and \textbf{arrays} require detailed 
representation in the abstract syntax structure from the table \ref{tab:node-attr}.

The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.

@@

\begin{description}
\item[structures] The structure types of B are naturally encoded as structure
  types in LLVM.

Declaring a structure variable and nested structure type:
%\[
%  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
%\] 
\[
\textcolor{gray}{T_1 =} \underbrace{struct( \underbrace{num1 \in \mathbb{Z}, num2 \in  \mathbb{Z} }_{seq\ Field} )}_{Struct};\\
  \textcolor{gray}{name1 \in} \ T_1;
\]

\[  
  \textcolor{gray}{name2 \in} \underbrace{struct( \underbrace{num1 \in \mathbb{Z}, elem \in  T_1 }_{seq\ Field} )}_{Struct}
\]
The simplified LLVM translation is:

\begin{llvmcode}
%struct.T_1 = type { i32, i32 }
%struct.T_2 = type { i32, %struct.T_1 }
%struct.state = type { %struct.T_1, %struct.T_2 }

@name2 =  global %struct.T_2 zeroinitializer
@name1 =  global %struct.T_1 zeroinitializer
\end{llvmcode}

The rules to declare the structure type are:
\begin{align*}
\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{TField}}} \AND\\
 l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
\end{align*}
$$\trad{t}{\B{TField}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$

The records types are initialized in the initializing function of the module.\\

\begin{comment}
The following rule is invoked in the initializing function of the module.
\begin{align*}
\trad{instanceModule,Structure}{\B{InitDerivedType}} \isdef \\
\LET name \BE \Global(\B{instanceModule,Structure.elements}) \AND\\
tlabel \BE \local(Structure.elements) \IN\\
\llvm{"\PH{name} = global \PH{tlabel} zeroinitializer\nl"}\\
\end{align*}
\end{comment}

\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.
Declaring an array type:
\[
\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..1) \times (0..1) \times (0..3)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
\]
\end{description}

Declaring an array: the essential elements of the array are its name, size and type. 
The following code declares a multidimensional global array called ``Arr'' 
with 16 (2*2*4) integer numbers with 32 bits.

\begin{llvmcode}
%struct.state = type { ... , [2 x [2 x [4 x i32]]],.. }
\end{llvmcode}

The function ArrayType returns the type for a declared variable array in 
structure state, for example:

\begin{align*}
\trad{Array}{\B{Type}} \isdef \\
& \trad{Array.dom , Array.ran.type}{ArrayType}\\
\trad{seqSet,type}{\B{ArrayType}} \isdef \\
& \LET h \BE \B{head}(seqSet) \AND\\
& times \BE h.end - h.start + 1 \IN\\
& \IF seqSet.length == 1 \THEN\\
& \llvm{"[\PH{times} x \PH{type}] \nl"} \\
& \ELSE \\
& \LET res \BE \trad{tail(seqSet),type}{ArrayType} \IN \\
& \llvm{"[\PH{times} x \PH{res}] \nl"} \\
\end{align*}

[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]

@@

To facilitate the definition of rules that apply to similar cases, we introduce
mapping $\opmap$, that associates B operators to corresponding LLVM keywords:
$$\opmap = \{
\begin{array}[t]{l}
  = \mapsto \llvm{"eq"},
  \ne \mapsto \llvm{"ne"},
  < \mapsto \llvm{"slt"},
  \le \mapsto \llvm{"sle"},
  > \mapsto \llvm{"sgt"},
  \ge \mapsto \llvm{"sge"}, \\
  + \mapsto \llvm{"add"} \}.
  \end{array}
$$
